---
title: "Power simulation for waterlogging experiment"
author: "Hamish McLean"
date: '2022-02-03'
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(ggpubr)
library(abind) # bind matrices
library(manipulate)
library(tictoc)
```

# Power simulation for waterlogging experiment

The purpose of this notebook is to simulate the waterlogging experiment and calculate power for different experimental designs.

## The experiment

Apple rootstocks will be planted in pots and arranged in a randomised complete block experimental design. A range of waterlogging treatments will be applied. Trees will be inoculated with *Neonectria ditissima* and canker symptoms will be assessed. Canker incidence will be compared for the different waterlogging treatments. 300 rootstocks will be available, which will be planted in threes in pots. Three inoculation points per tree will ensure there is no interaction between lesions. Pots will be grouped together in plots for more inoculation points per plot.

## Parameters

```{r}
trees <- 300
treesPerPot <- 3

potsPerPlot <- 4
treesPerPlot <- treesPerPot * potsPerPlot
plots <- trees %/% treesPerPlot

treatments <- 5
blocks <- plots %/% treatments
remainder <- trees - blocks * treatments * treesPerPlot

pointsPerTree <- 3
pointsPerPlot <- pointsPerTree * treesPerPlot

infectionRate <- 0.15
effect <- 1.5


data.table(
    factor = c("treesPerPlot", "plots", "treatments", "blocks", "remainder", 
               "infectionRate", "effect"),
    value = c(treesPerPlot, plots, treatments, blocks, remainder, 
              infectionRate, effect)
)
```

## Single experiment

For a single iteration of the experiment, canker infections are randomly sampled from a binomial distribution, $y\sim{}Bin(p, n)$, with probability estimated as `infectionRate` and n as the number of inoculation points per plot as `pointsPerPlot`.

```{r}
singleExperiment <- function(treatment, blocks, pointsPerPlot, infectionRate){
    treatment <- rep(treatment, blocks)
    incidence <- rbinom(blocks, pointsPerPlot, infectionRate)/pointsPerPlot
    n <- rep(pointsPerPlot, blocks)
    data.table(treatment, incidence, n)
}

singleData <- rbind(
  singleExperiment("control", blocks, pointsPerPlot, infectionRate),
  singleExperiment("waterlogged", blocks, pointsPerPlot, infectionRate * effect)
)

singleModel <- glm(incidence ~ treatment, family = binomial(link = 'logit'), 
    data = singleData, weights = n)

summary(singleModel)
```

```{r}
gghistogram(data = singleData, 
          x = "incidence",
          add = "mean",
          binwidth = 0.05,
          color = "treatment", fill = "treatment",
          position = "dodge")
          #palette = c("#00AFBB", "#E7B800"))

ggstripchart(data = singleData, 
             x = "treatment",
             y = "incidence",
             add = "mean_ci",
             color = "treatment",
             add.params = list(color = "black")) #+
  #stat_compare_means(method = 'kruskal.test')#comparison = list(c("control", "waterlogged")))

```

## Power calculation

To calculate power, the simulation must be repeated many times. Power is calculated as the proportion of p values less than alpha (0.05 by default) after fitting each repeat to a glm. For this implementation, a matrix (with `blocks` rows and `iterations` columns) of random values sampled from a binomial distribution with each column as a single iteration of the experiment. A glm is fit to each column and the p values recorded.

```{r}
#' Simulate power by generating data from a binomial distribution,
#' fitting to a binomial glm and calculating power.
simulatePower <- function(iterations, blocks, points, rate, effect, alpha=0.05){
    
    # Generate data
    control <- replicate(iterations, rbinom(blocks, points, rate)/points)
    waterlogged <- replicate(iterations, 
                             rbinom(blocks, points, rate * effect)/points)
    data <- abind(control, waterlogged, along = 1)
    
    # Fit model and calculate p values
    treatment <- c(rep("control", blocks), rep("waterlogged", blocks))
    n <- rep(points, 2 * blocks)
    pValues <- apply(data, 2, function(y) calculateP(y, treatment, n))
    
    # Calculate power
    sum(pValues <= alpha) / length(pValues)
}

#' Fit a binomial glm and return the p value.
calculateP <- function(y, x, weights){
    model <- glm(y ~ x, family = binomial(link = 'logit'), weights = weights)
    coef(summary(model))[2,4]  # return p value from model
}

#simulatePower(iterations = 1000, blocks, pointsPerPlot, infectionRate, effect)
```

## Varying parameters

To compare power for different experimental designs, the experimental parameters are varied according to different designs. For this, a table is generated from every combination of the variable parameters. The power simulation is then run for each combination of parameters.

```{r}
trees <- 300
treesPerPot <- 3
pointsPerTree <- 3

potsPerPlot <- 4
treatments <- c(3, 5, 7)
infectionRate <- c(0.1, 0.15, 0.2)
effect <- c(1.4, 1.7, 2)

experiments <- expand.grid(
    potsPerPlot = potsPerPlot,
    treatments = treatments,
    rate = infectionRate, 
    effect = effect
) %>%
    as.data.table() %>% 
    .[, treesPerPlot := treesPerPot * potsPerPlot] %>% 
    .[, plots := trees %/% treesPerPlot] %>% 
    .[, blocks := plots %/% treatments] %>% 
    .[, points := pointsPerTree * treesPerPlot] %>% 
    .[, remainder := trees - blocks * treatments * treesPerPlot]

```

```{r}
iterations <- 5000

#' Simulate power for each combination of parameters in `experiments`.
#' This function takes a long time to run!
powerComparison <- function(iterations, experiments){
    experiments[, power := mapply(simulatePower, iterations, 
                                  blocks, points, rate, effect)]
}

# This takes a long time to run!
tic(paste(nrow(experiments), "simulations of", iterations, "iterations"))
powers <- powerComparison(iterations, experiments)
toc()

#View(powers)
```

## Results

```{r}
ggplot(powers, aes(effect, power, group = rate, 
                 color = cut(rate, breaks = 3, labels = infectionRate))) +
    geom_point() +
    geom_line() +
    geom_hline(yintercept = 0.8, linetype = "dotted", col = "red") +
    scale_y_continuous(breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), limits = c(0, 1)) +
    #facet_grid(potsPerPlot ~ treatments, labeller = label_both) +
    facet_wrap(~treatments, labeller = label_both) +
    scale_color_viridis_d(name = "Infection Rate") +
    theme_bw() + 
    ggtitle(paste("Power simulation with", iterations, "iterations"))
```

`infectionRate` and `effect` size has a large effect on power. This makes it difficult to reliably predict power as the values of these variables must be estimated.

`potsPerPlot` has little effect on power. More `potsPerPlot` results in more inoculation points per plot (`pointsPerPlot`), but fewer blocks and vice versa.

`treatments` has a large effect on power. 3 treatments (control and 2 waterlogging treatments) has sufficient power for effect sizes above 1.6. 5 treatments (control and 4 waterlogging treatments) only has sufficient power for effect sizes above 1.7 and only with `infectionRate` of 0.15 or above.
