---
title: "Precipitation analysis"
author: "Hamish McLean"
date: "01/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# data aquisition
library(RCurl)
library(readr)

# data wrangling
library(here)
library(data.table)
library(magrittr)

# time series
library(xts)
library(tsbox)
library(zoo)  # rolling mean

# graphing
library(ggplot2)
library(ggsci)  # color palettes
library(ggpubr)  # publication quality plots
library(ggforce)  # better jitter
library(Amelia)  # missmap
library(gridExtra)  # grids

library(hydroTSM)  # precipitation data

here <- here::here
```

## Data

UK hourly rainfall data for East Malling, Kent was downloaded from: <https://data.ceda.ac.uk/badc/ukmo-midas-open/data/uk-hourly-rain-obs/dataset-version-202107/kent/00744_east-malling/qc-version-1>

The table description can be found here: <https://artefacts.ceda.ac.uk/badc_datadocs/ukmo-midas/RH_Table.html>

```{r warning=FALSE}
dataFolder <- "data/hourly-rain_east-malling"
filePrefix <- "hourly-rain_east-malling_"

startYear <- 2000
endYear <- 2020
years <- startYear:endYear

columns <- c("ob_end_time", "met_domain_name", "prcp_amt")

files <- lapply(years, function(year) here(dataFolder, 
                                           paste0(filePrefix, year, ".csv")))

data <- rbindlist(lapply(files, fread, skip = 61))#, select = columns))

#View(data)
```

Data includes hourly and daily observations `data$met_domain_name` AWSHRLY and AWSDLY. Hourly observations (AWSHRLY) are recorded on the hour according to `data$ob_end_time`. Daily observations (AWSDLY) are recorded every 12 hours according to `data$ob_end_time` from T09-21Z and T21-09Z.

```{r}

awsdly <- data[met_domain_name == "AWSDLY"]

#View(awsdly)

summary(awsdly)

summary(awsdly[ob_hour_count > 100])

plot(awsdly$ob_end_time, awsdly$prcp_dur_q)

unique(awsdly$rec_st_ind)

awsdly[prcp_amt > 80]

unique(awsdly$prcp_amt_q)

awsdly$prcp_amt_q

grid.arrange(
  qplot(awsdly$ob_end_time, awsdly$ob_hour_count),
  qplot(awsdly$ob_end_time, awsdly$prcp_amt),
  qplot(awsdly$ob_end_time, awsdly$prcp_amt_q),
  qplot(awsdly$ob_end_time, awsdly$midas_stmp_etime)
)
```

Summary results

-   `ob_end_time` is the time of the end of the observation. This includes the full date range, but some values are missing.

-   `id` is the raingauge ID. This was 297881 for all observations.

-   `ob_hour_count` is observation period. This is mostly 12, but includes higher values. This is likely due to larger observation windows where daily observations are missing.

-   `version_num` is the QC version number. All values are 1.

-   `src_id` is the station ID. All values are 744 which is the East Malling weather station.

-   `rec_st_ind` relates to QC and includes the following values: 1001, 1011, 1014, 1022. These can be interpreted at <https://dap.ceda.ac.uk/badc/ukmo-midas/metadata/doc/state_indicators.html>

-   `prcp_amt` is mostly 0, but commonly ranges up to 40, some higher values (80 - 100) could be outliers. All values above 80 correspond with `ob_hour_count` of 744

-   `prcp_dur` is all NA

-   `prcp_amt_q` is the QC code for `prcp_amt` this highlights QC problems

-   `prcp_dur_q` is mostly NA with some 0s, not relevant because `prcp_dur` not used

-   `prcp_amt_j` is all NA

-   `meto_stmp_time` is the actual time data was received

-   `midas_stmp_etime` is the elapsed time to storage

Data cleaning

-   Remove `ob_hour_count` values above 24 for the daily data set. Keep values up to 744 (30 days) for the monthly data set.

-   Add all missing dates to `ob_hour_count` set values for all variables to NA.

-   Check for missing values and remove whole years or months with too many missing values.

-   Keep variables: `ob_hour_count` and `prcp_amt` and rename to `date` and `precipitation` .

```{r}

daily <- awsdly[ob_hour_count == 12]  # filter out longer observation periods

daily <- daily[, .(date = as.Date(ob_end_time), precipitation = prcp_amt)] %>%
  .[, .(precipitation = sum(precipitation)), by = date]  # sum day and night

#View(daily)
```

## Data validation

```{r}
startDate <- as.Date(paste0(start_year, "-01-01"))
endDate <- as.Date(paste0(end_year, "-12-31"))
dateRange <- data.table(date = seq(startDate, endDate, "days"))

daily <- daily[dateRange, on = "date"]  # Include missing dates as NA

daily[is.na(precipitation), .N, by = .(year(date), month(date))]

d <- dcast(daily, yday(date) ~ year(date))[]
missmap(d, rank.order = FALSE)
```

| Year | Month | NA  |
|------|-------|-----|
| 2006 | 08    | 30  |
| 2006 | 10    | 30  |
| 2007 | 06    | 29  |
| 2007 | 07    | 30  |
| 2007 | 08    | 30  |
| 2007 | 09    | 29  |
| 2007 | 10    | 30  |
| 2008 | 01    | 30  |

## Data cleaning

Delete whole of 2007 (just delete entirely)

Delete whole months with missing data (fill with NA)

-   2006-08

-   2006-10

-   2008-01

```{r}
daily <- daily[year(date) != 2007]
daily[date %between% .("2006-08-01", "2006-08-31") 
      | date %between% .("2006-10-01", "2006-10-31")
      | date %between% .("2008-01-01", "2008-01-31"), 2] <- NA

daily[is.na(precipitation), .N, by = .(year(date), month(date))]

d <- dcast(daily_data, yday(date) ~ year(date))[]
missmap(d, rank.order = FALSE)
```

## Monthly data

```{r}
monthly <- daily[, .(
    prcpMean = mean(precipitation, na.rm = TRUE), 
    prcpSum = sum(precipitation, na.rm = TRUE),
    prcpN = sum(!is.na(precipitation)),
    prcpPerc = sum(precipitation > 0) / sum(!is.na(precipitation)) * 100
), by = .(year(date), month(date))]

#View(monthly)

ggstripchart(data = monthly, 
             x = "month",
             y = "prcpMean",
             color = "year",
             add = "mean_se")

ggplot(monthly, aes(x = month, y = prcpMean)) + 
    stat_summary(geom = "ribbon", fun.data = mean_cl_normal, width = 0.1, 
                 conf.int = 0.95, fill = "lightblue") +
    stat_summary(geom = "line", fun = mean, linetype = "dashed") +
    stat_summary(geom = "point", fun = mean, color = "red") + 
    labs(x = "Month", y = "Mean daily precipitation (mm/day)", 
         title = paste0("Mean daily precipitation for ", startYear, '/', endYear)) +
    scale_x_discrete(limits = month.abb) + 
    theme_classic2()

ggplot(monthly, aes(x = month, y = prcpSum)) + 
    stat_summary(geom = "ribbon", fun.data = mean_cl_normal, width = 0.1, 
                 conf.int = 0.95, fill = "lightblue") +
    stat_summary(geom = "line", fun = mean, linetype = "dashed") +
    stat_summary(geom = "point", fun = mean, color = "red") + 
    labs(x = "Month", y = "Mean monthly precipitation (mm/month)", 
         title = paste0("Mean monthly precipitation for ", startYear, '/', endYear)) +
    scale_x_discrete(limits = month.abb) + 
    theme_classic2()

```

## Rain events

```{r}
rainEvents <- daily[, .(rainPerc = sum(precipitation > 0) / .N * 100, N = .N), 
                    by = .(month(date), year(date))] %>% 
    .[, monthName := factor(month.abb[month],levels=month.abb)]


ggplot(rainEvents, aes(x = month, y = rainPerc)) + 
    stat_summary(geom = "ribbon", fun.data = mean_cl_normal, 
                 width = 0.1, conf.int = 0.95, fill = "lightblue") +
    stat_summary(geom = "line", fun = mean, linetype = "dashed") +
    stat_summary(geom = "point", fun = mean, color = "red") +
    labs(x = "Month", y = "Rainy days (%)", 
         title = paste0("Percentage of rainy days per month for ", 
                        startYear, '/', endYear)) +
    scale_x_discrete(limits = month.abb) + 
    theme_classic2()

```

## Rolling mean

```{r}

daily <- daily[, .(date, precipitation, prcpRoll5 = frollmean(precipitation, 30))]
plot(daily$date, daily$prcpRoll5)

dailyYears <- daily[, .(precipitation = mean(precipitation, na.rm = TRUE)), 
                    by = .(doy = format(date, "%j"))] %>% 
  .[, .(doy, prcpRoll5 = frollmean(precipitation, 5))]

View(dailyYears)
plot(dailyYears$doy, dailyYears$prcpRoll5, type = "b")
```

## XTS

```{r fig.height=10}

xtsDaily <- as.xts(daily)
View(xtsDaily)

hydroplot(xtsDaily, var.type = "Precipitation", xlab = "date", ylab = "prcp")

periodicity(xtsDaily)

dwi(xtsDaily, out.unit="mpy")

m <- daily2monthly(xtsDaily, FUN=sum, na.rm=TRUE)
# Creating a matrix with monthly values per year in each column
M <- matrix(m, ncol=12, byrow=TRUE)
colnames(M) <- month.abb
rownames(M) <- unique(format(time(m), "%Y"))
# Plotting the monthly precipitation values
require(lattice)  # Loading required package: lattice
print(matrixplot(M, ColorRamp="Precipitation",
                 main="Monthly precipitation at East Malling (mm/month)"))


cmonth <- format(daily$date, "%b")  # Creating ordered monthly factors
months <- factor(cmonth, levels=unique(cmonth), ordered=TRUE)
# Boxplot of the monthly values
boxplot(coredata(xtsDaily) ~ months, col="lightblue", main="Monthly Precipitation",
ylab="Precipitation, [mm]", xlab="Month")


seasonalfunction(xtsDaily, FUN=sum, na.rm=TRUE) / length(years)
hydroplot(xtsDaily, pfreq="seasonal", FUN=sum, stype="default")



plot(ts_ts(xtsDaily))

decomposition = stl(ts_ts(xtsDaily), s.window=31)#, na.action = na.omit)

plot(decomposition)


monthly_rainfall <- period.apply(historical, INDEX = seq(1, nrow(historical) - 1, 30.4375), FUN = sum)

plot(ts_ts(monthly_rainfall))
```
